---
title: "Data Handling & ggplot"
author: "Toheeb"
date: "`r Sys.Date()`"
output: github_document
---

# **PART 1**

## Reading data
```{r}
carseats <- read.csv("G:/My Drive/Programming/Data Science/DATA SETS/Carseats.csv")
View(carseats) # displays the data in a separate tab in RStudio
```

```{r}
age <- carseats$Age
educ <- carseats$Education
age_educ <- cbind(age, educ)
write.table(age_educ, file = "G:/My Drive/Programming/Data Science/R Learning/Data Handling & Graphics/age_educ.csv") #this syntax stores the newly created data frame in specified format and location.
```



## Exploring Data
```{r}
my_data <- airquality # "airquality" is a data set readily available in R database.
```

```{r}
head(my_data) # prints the first 6 elements of the data
head(my_data, n = 4) # first 4 rows
head(my_data, n = -10) # all but the last 10
```

```{r}
tail(my_data)# last 6
tail(my_data, n = 12) # last 12
tail(my_data, n = -20) # all but the first 20
```


```{r}
class(my_data) # tells the class/type of the object (here it is a data frame)
length(my_data) # length of the object (i.e., no. of cols in the df)
ncol(my_data) # same as above
nrow(my_data) # no. of rows
names(my_data) # names of columns
rownames(my_data) # index names (rows)
str(my_data) # gives a brief overview of the df, including data type and entries
summary(my_data) # gives summary statistics (mean, median, min, etc.) for each column
```


```{r}
library("dplyr")
glimpse(my_data) # alternative to the str() function
```


```{r}
my_data$Month <- factor(my_data$Month) # converts the Month var to factor type (coz the var is categorical)
```

```{r}
glimpse(my_data)
levels(my_data$Month) <- list(May = "5", June = "6", July = "7", Aug = "8", Sept = "9") # labels the values/levels of the Month var
levels(my_data$Month) # lists the categories in the Month column
```





## Plotting data
```{r}
plot(my_data$Ozone, my_data$Temp)
idx <- identify(my_data$Ozone, my_data$Temp)
# this syntax allows to click on specific points on the graph.
```

The default labeling is row number. This can be changed as follows:
```{r}
plot(my_data$Ozone, my_data$Temp)
idx <- identify(my_data$Ozone, my_data$Temp, labels = my_data$Month, plot = TRUE) # labels the data points with the corresponding month.
```

Even better:
```{r}
plot(my_data$Ozone, my_data$Temp)
idx <- identify(my_data$Ozone, my_data$Temp, labels = paste(as.character(my_data$Day),
                      "-", as.character(my_data$Month)), plot = TRUE)
```


PS: all the codes above (on plot labeling) don't work in R Markdown. But they work in a normal R script (in a basic R environment)


Get the number of observations in each level/category of a var. For example, the variable Month:
```{r}
xtabs(~Month, my_data) 
```



## Dealing with NA's (missing values) 
NA indicates a missing case/obs \
NaN = "not a number" (e.g., in cases where a math operation is ran on a non-numeric var)

Although there already are missing cases in the current data set, one could throw in more:
```{r}
my_data[154, ] <- c(NA) # adds another row all with NA values
my_data[, 7] <- c(NA) # adds a new col all having NA values
View (my_data)
```

```{r}
View(is.na(my_data)) # checks for all NA values anywhere in the data set, returns table displaying TRUE/FALSE
```

```{r}
any(is.na(my_data)) # returns TRUE if at least one missing case exists in the data
```

```{r}
all(is.na(my_data)) # checks if all entries in the specified object (i.e., the entire dataset) are missing values
```

```{r}
all(is.na(my_data[,7])) # checks if all entries in the specified column are missing values
```


```{r}
# The recently added row and column contain only missing cases, so they can be removed:
my_data <- my_data[-7] # removes the last column (no. 7)
my_data <- my_data[-154, ] # removes the last row (no. 154)
```


```{r}
any(is.na(my_data)) # rechecks for missing cases
sum(is.na(my_data)) # gives the total no. of missing obs in the entire data set
```

```{r}
sum(is.na(my_data$Solar.R)) # gives the no. of missing obs in a column
```


```{r}
colSums(is.na(my_data)) # shows the distribution of missing cases across all columns
```

The rows of missing observations can be deleted as follows:
```{r}
clean_data <- na.omit(my_data) # generates a new data with all missing cases removed.
nrow(clean_data)
```

```{r}
nrow(my_data)-nrow(clean_data) # gives the difference between the no. of rows in the original data and the # rows in the cleaned data (i.e., no. of missing obs rows deleted)
```

Alternatively:
```{r}
clean_data2 <- my_data[complete.cases(my_data), ] # this code indexes only the rows with complete cases (no missing values)
nrow(clean_data2)
```

NB: These two approaches greatly reduce sample size. To preserve sample size, one may remove NA values only from variables to be used for analysis, e.g, if Ozone is not needed for analysis:
```{r}
clean_data3 <- na.omit(my_data[-1]) # this removes the NA values from all but col 1 (Ozone)
nrow(clean_data3)
```

```{r}
nrow(my_data)-nrow(clean_data3) # only 7 rows removed (as opposed to 42 rows deleted with the previous approach)
```

In addition, missing values can be removed based on some predefined condition(s):
```{r}
clean_data4 <- my_data[, colSums(is.na(my_data)) < 10] # a new data containing only columns with less than 10 missing cases
```

```{r}
final_data <- na.omit(clean_data4)
nrow(final_data)
```








## Combining data
```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS") # changes the working directory
getwd() #retrieves the current working directory
owl_morph <- read.csv("owl.morphometrics.csv", header = TRUE)
```

NB: In an R Markdown document, changing the working directory affects only the host chunk, not entire script. That is, the directory resets after that chunk resets. E.g., compare the getwd() output of the chunk above and that of the next chunk:
```{r}
getwd()
#retrieves the current working directory
```

```{r}
summary(owl_morph)
View(owl_morph)
any(is.na(owl_morph))
str(owl_morph)
```

```{r}
owl_morph$common.name <- as.factor(owl_morph$common.name)
owl_morph$latin <- as.factor(owl_morph$latin)
```

Both columns have now been changed to factor (i.e., categorical variables)

```{r}
plot(owl_morph$weight.g, owl_morph$wingspan.cm, xlab = "Owl weight (g)", ylab = "Owl wingspan (cm)")
```


```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS")
owl_clutch <- read.csv("owl.clutch.size.csv", header = TRUE)
View(owl_clutch)
any(is.na(owl_clutch))
```

Since the two datasets have the same contents in row 1 (common.name) orderly arranged, they can be combined using the cbind function:
```{r}
owl_morph_clutch <- cbind(owl_morph, owl_clutch)
View(owl_morph_clutch)
```

NB: The combination syntax above leads to duplication of the "common name row". This can be avoided:
```{r}
owl_morph_clutch <- cbind(owl_morph, owl_clutch[, 2]) # only the row 2 of the second data is called upon for the combination
```

```{r}
names(owl_morph_clutch)[6] <- "clutch.size" # renames the newly added column
```

```{r}
plot(owl_morph_clutch$wingspan.cm, owl_morph_clutch$clutch.size,
     xlab = "Owl wing span (cm)", ylab = "Owl clutch size (??)")
```


```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS")
owl_lspan <- read.csv("owl.lifespan.csv", header = TRUE)
View(owl_lspan)
any(is.na(owl_lspan))
```

This new data has one NA, and the identifying row (common.name) is not arranged alphabetically. To combine it with the recent combined data, use the merge() function and specify the column to be used for combination
```{r}
owl_morph_clutch_life <- merge(owl_morph_clutch, owl_lspan, key = "common.name")
View(owl_morph_clutch_life)
```


# **PART 2**

## The apply() Family of Functions 

### 1. apply() 
It has the general structure: apply(object, margin, function...); the object is usually a matrix or an array.\
Consider the ffg example:
```{r}
duckweed_mat <- matrix(c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 10, 30, 50, 80, 100, 150, 200,
                         250, 270, 300, 10, 30, 36, 80, 96, 106, 110, 130, 136, 144, 10, 15, 30,
                         50, 70, 86, 95, 100, 105, 190, 10, 40, 50, 65, 78, 96, 107, 120, 144,
                         157, 10, 30, 57, 98, 106, 130, 160, 177, 189, 198),
                       nrow = 10, byrow = FALSE)
rownames(duckweed_mat) <- c("Day1", "Day2", "Day3", "Day4", "Day5", "Day6", "Day7", 
                            "Day8", "Day9", "Day10")

colnames(duckweed_mat) <- c("R1", "R2", "R3", "R4", "R5", "R6")
duckweed_mat
class(duckweed_mat)
```

Say we want to check the maximum number of leaves recorded each day, i,e, per row:
```{r}
max(duckweed_mat[1,])
max(duckweed_mat[2,])
max(duckweed_mat[3,])
```
...and so on.\
Alternatively, one could use a FOR loop:
```{r}
for (i in 1:10) {
  row <- duckweed_mat[i, ]
  max <- max(row)
  print(max)
}
```

All this is done better with the apply() function:
```{r}
apply(duckweed_mat, 1, max) # returns the maximum value in each row
#NB: the 2nd argument is the margin -- 1 represents row; 2 implies column
```

Could do the same column-wise
```{r}
apply(duckweed_mat, 2, max)
# this prints out the maximum value in each column
```

The apply() function also works for data frames:
```{r}
duckweed_df <- data.frame(R1 = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
                          R2 = c(10, 30, 50, 80, 100, 150, 200, 250, 270, 300),
                          R3 = c(10, 30, 36, 80, 96, 106, 110, 130, 136, 144),
                          R4 = c(10, 15, 30, 50, 70, 86, 95, 100, 105, 190),
                          R5 = c(10, 40, 50, 65, 78, 96, 107, 120, 144, 157),
                          R6 = c(10, 30, 57, 98, 106, 130, 160, 177, 189, 198))
class(duckweed_df)
```
To calculate the means by row
```{r}
rowMeans(duckweed_df)
```

Using the apply() function:
```{r}
apply(duckweed_df, 1, mean)
```

However, unlike matrices and arrays, a data frame can contain non-numeric elements:
```{r}
duckweed_df$Day <- as.factor(1:10)
(duckweed_df)
```

NB: although the newly added column contains values 1 to 10, these values have been assigned as non-numeric (factor)
```{r}
duckweed_df <- duckweed_df[, c(7, 1:6)] # rearranges the data so that the new column comes 1st
class(duckweed_df$Day) # check that the column is correctly identified as factor 
```

Now using the apply function again:
```{r}
apply(duckweed_df, 1, mean)
```

NB: the above returns an error message ("Warning: argument is not numeric or logical: returning NA") because each row now has a factor, a data type that is not receptive to arithmetic operations.
```{r}
apply(duckweed_df[, 2:7], 1, mean) # the non-numeric column is now excluded
apply(duckweed_df[, -1], 1, mean) # can also be excluded this way
apply(duckweed_df[, -c(1,2,4,6)], 1, mean) # can also exclude several columns
```

Throw a missing observation into the mix:
```{r}
duckweed_df[6, 5] <- NA
(duckweed_df)
apply(duckweed_df[, -1], 1, mean)
```

NB: Now that there is a missing value in a row, that row returns a mean of "NA". If that specific cell is not of interest, one could instruct R to ignore it and compute the mean for only the non-missing values in that row:
```{r}
apply(duckweed_df[, -1], 1, mean, na.rm = TRUE) # na.rm "removes" the missing value(s) from the calculation
```

#### Using customized functions
The examples thus far have used a built-in R function (i.e., mean). But customized functions can also be used. For example, one may be interested in calculating, for each plant, the no. of leaves counted each day as a  proportion of the total no. of leaves (i.e, the count on the 10th day) for that plant:
```{r}
prop <- function(x) {
  x / max(x)
}
```

Using the matrix data:
```{r}
(duckweed_mat)
apply(duckweed_mat, 2, prop)
```

Note that the calculation is per column, hence the margin argument is set to "2". \
Basically, the apply() function takes each column as a vector and each cell as an element in that vector.

### 2. lapply()
General structure: lapply(object, function, ...) \
-> the "object" could be vector, list, or data frame. \
-> lapply() returns only list outputs. \
Example - Below is an hypothetical data on the clutch sizes of Canada geese under 4 different diet regimes, with 10 replicates in each diet group:
```{r}
cago_list <- list(Diet1 = c(2, 5, 4, 5, 3, 5, 4, 4, 4, 5),
                  Diet2 = c(8, 5, 6, 5, 7, 7, 6, 8, 8, 3),
                  Diet3 = c(3, 4, 2, 5, 2, 6, 5, 6, 2, 4),
                  Diet4 = c(2, 2, 3, 2, 5, 2, 4, 3, 5, 7))
class(cago_list)
(cago_list)
```

Use the lapply() function on the data:
```{r}
lapply(cago_list, mean)
```

NB: each row is a vector, and the mean is thus returned per vector (i.e., diet).

The same data could be stored in a data frame instead:
```{r}
cago_df <- data.frame(Diet1 = c(2, 5, 4, 5, 3, 5, 4, 4, 4, 5),
                  Diet2 = c(8, 5, 6, 5, 7, 7, 6, 8, 8, 3),
                  Diet3 = c(3, 4, 2, 5, 2, 6, 5, 6, 2, 4),
                  Diet4 = c(2, 2, 3, 2, 5, 2, 4, 3, 5, 7))
lapply(cago_df, mean)
```

Again, using the lapply returns the output in a list format. There is also no margin argument, unlike apply(). \

A vector example:
```{r}
random <- c("This", "is", "a", "random", "vector")
```

The no. of letters in each of the words contained in the vector can be determined as follows:
```{r}
lapply(random, nchar)
```

NB -- unlike the lapply() function, the following syntax returns an output in a vector format:
```{r}
nchar(random)
```



### 3. sapply()
--> It has the same general structure as lapply() and also applies to lists, vectors, and data frames. However, outputs are returned in a simplified format if possible (vector, matrix, or list). In addition, as in lapply(), sapply() has no margin argument.
```{r}
cago_list
sapply(cago_list, mean) # results are in a vector format
```

```{r}
cago_df
sapply(cago_df, mean) # results also in vector format
```

```{r}
random
sapply(random, nchar)
```

Sometimes, the output could not be simplified to a vector or matrix:
```{r}
sequence <- function(x) {
  seq(nchar(x))
}
```

As a reminder, the seq() function returns a sequence from 1 to number of specified element. See ffg examples:
```{r}
seq(3) # this outputs "1 2 3"
seq(nchar("that")) #this outputs "1 2 3 4"
```
Now, apply the created function above
```{r}
sapply(random, sequence)
```

As displayed, the output is not simplified as vector.


### 4. tapply() 
General structure: tapply(x, index, function, ...) \
--> It applies to object subsets (vector, column of a data frame, element of a list, etc...)
```{r}
patient_ID <- 1:30
age <- c(32, 45, 44, 34, 23, 26, 37, 45, 12, 23, 44, 35, 57, 65, 76, 43, 42, 34,
         36, 37, 23, 21, 28, 24, 29, 13, 18, 32, 25, 28)
treatment <- c("a", "c", "c", "b", "b", "b", "c", "b", "c", "a", "a", "a", "a", "a","b",
               "b", "b", "b", "c", "c", "c", "a", "b", "c", "b", "a", "a", "c", "a", "c")
```

```{r}
tapply(age, treatment, mean) # gives the mean age by treatment groups
```

Store the data in a data frame, and apply the function:
```{r}
med_df <- data.frame(patient_ID, treatment, age)
head(med_df) # Recall: 6 rows by default (same as the tail fxn)
```

```{r}
tapply(med_df$age, med_df$treatment, mean)
```

A list could also work:
```{r}
med_list <- list(patient_ID = patient_ID, treatment = treatment, age = age)
tapply(med_list$age, med_list$treatment, mean)
```


# **PART 3**

## aggregate() function 
--> Similar to tapply() but with wider applications \
--> General structure: aggregate(formula, data, function, ...) \
--> works with list or data frame (other data structures will be converted to either of these two)


Example -- Using the co2_uptake data (a built-in data set in R that contains an experiment data of CO2 uptake in plants from two different locations):
```{r}
co2_uptake <- CO2
head(co2_uptake)
```

Add a new column to the data:
```{r}
co2_uptake$height <- c(35.77, 43.95, 38.10, 43.20, 43.02, 39.19, 31.60, 36.88, 41.11, 43.64,
                       36.82, 33.86, 30.17, 36.92, 36.15, 43.60, 32.35, 43.92, 40.50, 37.46, 
                       33.92, 42.19, 30.20, 35.64, 39.63, 36.39, 42.95, 33.88, 43.75, 41.10, 
                       34.57, 30.21, 37.19, 33.45, 40.93, 32.93, 36.21, 40.74, 32.87, 35.98,
                       43.57, 39.91, 35.02, 33.20, 37.89, 34.96, 30.99, 40.12, 33.33, 34.48,
                       38.22, 35.21, 39.60, 40.29, 42.90, 36.09, 38.75, 36.65, 31.51, 39.32, 
                       30.27, 34.21, 40.59, 43.67, 32.10, 30.08, 42.10, 36.60, 43.89, 38.33,
                       36.99, 32.05, 37.54, 34.51, 33.69, 41.80, 30.91, 39.23, 30.93, 42.73,
                       34.82, 33.20, 31.57, 43.32)

View(co2_uptake)
```

There are 7 concentration categories (95,175,250,350,500,675,1000) in the data. To calculate, say, mean uptake for each conc category:
```{r}
aggregate(uptake~conc, co2_uptake, mean) # tabular output
```

As already mentioned, the same operation could be performed using tapply(), but the output is in a vector format (unlike the aggregate fxn which produced a tabular output):
```{r}
tapply(co2_uptake$uptake, co2_uptake$conc, mean) # vector output
```

Even better, aggregate() can handle multiple subsetting. For example, to calculate the mean uptake by concentration category and treatment group:
```{r}
aggregate(uptake ~ conc + Treatment, co2_uptake, mean)
```

...or, mean uptake by conc, treatment, and type:
```{r}
aggregate(uptake ~ conc + Treatment + Type, co2_uptake, mean)
```

Operations can also be performed on multiple elements/columns. For example, to calculate the mean uptake and mean height by conc:
```{r}
aggregate(cbind(uptake, height) ~ conc, co2_uptake, mean)
```


aggregate() also works on lists:
```{r}
co2uptake_list <- list(plant = co2_uptake$Plant, type = co2_uptake$Type, 
                       treatment = co2_uptake$Treatment, conc = co2_uptake$conc, 
                       uptake = co2_uptake$uptake, height = co2_uptake$height)
```


```{r}
aggregate(cbind(uptake, height) ~ conc, co2uptake_list, mean)
```

NB: the output structure is the same as for data frame. \


Other functions can also be applied using aggregate(). For example:
```{r}
aggregate(uptake ~ conc, co2uptake_list, length) # no. of uptake values recorded for each conc group
aggregate(height ~ conc, co2uptake_list, length) # no. of height values per conc group
aggregate(type ~ conc, co2uptake_list, length) # no. of type values per conc group
```




## paste() function
This function allows to merge multiple elements together as one. For example, working with a sample of birds, one could assign each individual a label containing their name and ID:
```{r}
paste("RWBL", 1)
```

Alternatively, the entries can be first stored before calling the paste() function:
```{r}
species <- "RWBL"
num <- 1
paste(species, num)
```

A "sep" argument can be added to define the separator (default is space):
```{r}
paste(species, num, sep = "-")
```

When paste() is used on a vector containing several elements, it doesn't work (i.e., each element in that vector maintains its independence), unless the "collapse" argument is properly invoked/defined:
```{r}
id <- c("RWBL", 1)
paste(id) # each element in "id" is pasted separately
```

```{r}
paste(id, collapse = " ") # default value for collapse is NULL
# the paste function now works as expected
```

```{r}
paste(id, collapse = "_")
```

NB: the "collapse" argument works like the "sep" argument; so it could take different values like space, underscore, period, etc. The default value is NULL. \

```{r}

```

The paste() function is really powerful when used with several vectors:
```{r}
species <- c("RWBL", "MODO", "AMRO", "AMCR", "MODO")
num <- 1:5
paste(species, num, sep = "_")
```

It the pasted vectors are of different length, the longer one starts recycling with the rest, E.g.:
```{r}
species <- c("RWBL", "MODO", "AMRO", "AMCR", "MODO")
num <- 1:8
paste(species, num, sep = "_")
```

This "recycling" feature comes in handy in the case of repetitions:
```{r}
num <- 1:10
paste("Bird", num) # the "Bird" element gets recycled through the num var, thus printing Bird 1, Bird 2, ... Bird 10.
```

Data frame application:
```{r}
df <- data.frame(species, num = 1:5)
paste(df$species, df$num, sep = "_") # as usual, the output is vector
```

The output of the paste() function can be saved in an object and put to further use:
```{r}
output = paste(df$species, df$num, sep = "_")
df$idnum = output
```

```{r}
View(df)
```



# **PART 4**

## Basic Plotting
```{r}
maxd <- 10
storks <- numeric(maxd)
babies <- numeric(maxd)
set.seed(15) # to retain the results of the rnorm functions
for (i in 1:maxd) {
  storks[i] <- rnorm(1, i * 100, 10) # Recall -- general form is rnorm(n, mean, sd)
  babies[i] <- rnorm(1, i * 10, 10)
}
```

```{r}
(storks)
(babies)
```

```{r}
plot(storks, babies)
abline(lm(babies ~ storks)) # fits a regression line to the plot

# let's plot a new graph on the existing plot: 
newstorkdata <- c(220, 411, 630, 705, 729, 850)
newbabydata <- c(11, 22, 38, 41, 45, 52)
points(newstorkdata, newbabydata, col = "red", pch = 19) #plots this on the existing graph; red  here is the color of the new points/observations
abline(lm(newbabydata ~ newstorkdata), col = "red") # red here is the color of the new line

# add a legend to the graph:
legend(100, 90, c("old data", "new data"), pch = c(1, 19), col = c("black", "red"))
# NB: The first argument for the legend function is "x,y". Here, the value is 100,90, i.e., the legend should be placed at the point on the graph where x=100 and y=90
```

```{r}

```

NB: using the locator function (as shown in the following code) works in basic R script, but not in R markdown: \
*legend(locator(1), c("old data", "new data"), pch = c(1, 19), col = c("black", "red"))*

One could also split the plot area as follows:
```{r}
par(mfrow= c(1, 2)) # splits the plot area into a row with 2 columns
# Recall that this code does not affect subsequent code chunks. However, in R scripts, the plot area must be reset to return to the original, non-split type, i.e., by running the code: par(mfrow = c(1,1))
```

The par() function can also be used to further define certain attributes of the graph(s), e.g.,: \
par(las = 1, cex = 1, cex.lab =1.2, cex.axis = 1.1) \
>-- las defines the style of axis labels (0=default=parallel, 1=horizontal, ...) \
>-- cex magnifies plotting texts and symbols relative to the current size \
>-- cex.lab magnifies axis labels relative to the current size \
>-- cex.axis magnifies axis annotations relative to the current size

## Creating box plots using R's ggplot2 package
```{r}
mydata <- ToothGrowth # ToothGrowth is a built-in data set in R
mydata$dose <- as.factor(mydata$dose) #changes the dose var to factor data type
View(mydata)
```

```{r}
# install.packages("tidyverse") 
library("ggplot2")
names(mydata)
```

Now, let's create a box plot:
```{r}
ggplot(mydata, aes(supp, len)) # creates an empty plot 
```

```{r}
ggplot(mydata, aes(supp, len)) + geom_boxplot() # inserts a boxplot into the empty plot created above
```

```{r}
# differentiate columns of box plots by the dose category
ggplot(mydata, aes(supp, len)) + geom_boxplot() + facet_grid(. ~ dose)
```

Alternatively, this "subsetting" can be done in the aesthetic ("aes") argument:
```{r}
ggplot(mydata, aes(supp, len, fill = dose)) + geom_boxplot() # this creates all box plots on a single graph, each category's plot differentiated by a distinct color.
```

Further, the color can be set such that darker/lighter color indicates an increase or decrease in levels of, for example, dose:
```{r}
my_color <- c("#fff7bc", "#fec44f", "#d95f0e") # yellow, orange, and red colors, respectively
ggplot(mydata, aes(supp, len, fill = dose)) + geom_boxplot() + scale_fill_manual(
                                    name = "Dose (mm)", breaks = c("0.5", "1", "2"),
                                   labels = c("Low 0.5mm", "Medium 1mm", "High 2mm"),
                                   values = my_color)
```

Change the background color to black and white:
```{r}
ggplot(mydata, aes(supp, len, fill = dose)) + geom_boxplot() + scale_fill_manual(
  name = "Dose (mm)", breaks = c("0.5", "1", "2"),
  labels = c("Low 0.5mm", "Medium 1mm", "High 2mm"),
  values = my_color) + theme_bw() # NB: the previous color was grey-ish
```


And a lot more...
```{r}
ggplot(mydata, aes(supp, len, fill = dose)) + geom_boxplot() + scale_fill_manual(
  name = "Dose (mm)", breaks = c("0.5", "1", "2"),
  labels = c("Low 0.5mm", "Medium 1mm", "High 2mm"),
  values = my_color) + theme(plot.background = element_blank(),
                             panel.grid.major = element_blank(),
                             panel.grid.minor = element_blank(),
                             panel.background = element_blank(),
                             axis.line = element_line(colour = "black"),
                             axis.title.x = element_text(size = 19, colour = "black"),
                             axis.title.y = element_text(size = 19, colour = "black"),
                             axis.text.x = element_text(size = 12, colour = "black"),
                             axis.text.y = element_text(size = 12, colour = "black"),
                             legend.text = element_text(size = 12),
                             legend.title = element_text(size = 15),
                             legend.background = element_rect(color = "black", 
                                                              size = .5,
                                                              linetype = "solid"),
                             legend.position = c(.60, .85)) + labs(x = "Supplement regimen",
                                                                   y = "Tooth length (mm)")
```




## Creating Histograms and density plots using ggplot2 
```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS")
getwd()
mydata <- read.csv("Mammal_lifehistories_v2.csv")
names(mydata)
View(mydata)
head(mydata)
summary(mydata)
ncol(mydata)
nrow(mydata)
```

Some of the cells have "-999" representing NA's (missing cases); remove such obs:
```{r}
mydata <- mydata[mydata$gestation.mo > 0 & mydata$mass.g > 0 & mydata$litter.size > 0, ]
```

```{r}
m <- ggplot(mydata, aes(x = gestation.mo))
m + geom_histogram()
```

Change the width of the bars:
```{r}
m + geom_histogram(binwidth = 0.4)
```

To create different histograms by, say, order categories (1st column in the current data):
```{r}
m + geom_histogram(binwidth = 0.4) + facet_grid(order ~ .)
```
NB: the result is too tiny to read because there are too many order categories, some well-sampled and others under-sampled.

To get the # of obs in each order category:
```{r}
xtabs(~ order, mydata)
```

Subset orders with enough obs (say at least 100 each):
```{r}
mydata_largeorders <- mydata[mydata$order == "Artiodactyla" | mydata$order == "Carnivora" |
                               mydata$order == "Primates" | mydata$order == "Rodentia", ]
```

Now rerun the operation on the subset data:
```{r}
m2 <- ggplot(mydata_largeorders, aes(x = gestation.mo))
m2 + geom_histogram(binwidth = 0.7) + facet_grid(order ~ .)
```



## Creating scatterplots using ggplot2 
```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS")
co2data <- read.csv("canada.co2.by.sector.csv")
names(co2data)
nrow(co2data); dim(co2data)
summary(co2data)
View(co2data)
head(co2data)

```

Remove rows/obs where the sector value is "Total, all sectors"
```{r}
co2data <- co2data[co2data$SECTOR != "Total, all sectors", ]
```

Create a scatterplot of output (in kilotonnes; column 4) vs. year:
```{r}
# install.packages("tidyverse") # package already installed
# library(ggplot2) # library opened already
co2_scatter <- ggplot(co2data, aes(year, log(co2_ann_kilotonnes), colour = 
                                     SECTOR)) + geom_point() + stat_smooth(method = "lm")
co2_scatter
```

The resulting scatterplot have some portions cut off because there are too many SECTOR categories. Try a subset of these categories:
```{r}
library("dplyr")
co2data_order <- co2data %>% group_by(SECTOR) %>% summarise(co2mean = 
                            mean(co2_ann_kilotonnes)) %>% arrange(desc(co2mean))
```
In the code chunk above: the co2data is grouped by the variable SECTOR, then the mean co2 ouptut of each sector is aggregated (displayed) in a single row then the rows are arranged by descending order of the means.
```{r}
View(co2data_order)
head(co2data_order)
```


```{r}
co2data_order_top10 <- co2data_order[1:10, ] # a new data set containing only the top 10 SECTORS by mean co2 output
```


```{r}
top10 <- c(co2data_order_top10$SECTOR) # concatenates the values of the SECTOR column
top10[5] # the 5th element in the vector
```

From the original data, cull out observations in the top 10 sectors into a new data:
```{r}
co2_top10 <- co2data[co2data$SECTOR == top10[1] | co2data$SECTOR == top10[2] |
                       co2data$SECTOR == top10[3] | co2data$SECTOR == top10[4] |
                       co2data$SECTOR == top10[5] | co2data$SECTOR == top10[6] |
                       co2data$SECTOR == top10[7] | co2data$SECTOR == top10[8] |
                       co2data$SECTOR == top10[9] | co2data$SECTOR == top10[10], ]
nrow(co2_top10)
tail(co2_top10)
View(co2_top10)
```

Create a scatterplot for the top 10 sectors:
```{r}
top10_scatter <- ggplot(co2_top10, aes(year, log(co2_ann_kilotonnes), colour = 
                                     as.factor(SECTOR))) + geom_point() + stat_smooth(method = "lm")
(top10_scatter)
```

```{r}
top10_scatter + theme_bw() # changes the background theme to black & white
```


Make further changes to the plot:
```{r}
top10_scatter <- ggplot(co2_top10, aes(year, log(co2_ann_kilotonnes), colour = 
                as.factor(SECTOR))) + geom_point() + stat_smooth(method = "lm") + theme(
                  plot.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  panel.background = element_blank(),
                  axis.line = element_line(colour = "black"),
                  axis.title.x = element_text(size = 19, colour = "black"),
                  axis.title.y = element_text(size = 19, colour = "black"),
                  axis.text.x = element_text(size = 12, colour = "black"),
                  axis.text.y = element_text(size = 12, colour = "black"),
                  legend.text = element_text(size = 12),
                  legend.title = element_text(size = 15),
                  legend.background = element_rect(color = "black", 
                                       size = .5,
                                       linetype = "solid")) + labs(x = "Year",
                    y = "Annual CO2 production (Log kilotonnes)") + scale_color_brewer(
                      palette = "Paired", name = "Sector")
top10_scatter
```





## Insetting Graphical Panels using ggplot2
```{r}
setwd("G:/My Drive/Programming/Data Science/DATA SETS")
mydata <- read.csv("Mammal_lifehistories_v2.csv")
names(mydata)
```

Remove observations with negative values (the coding for missing cases in the current data):
```{r}
mydata <- mydata[mydata$gestation.mo > 0 & mydata$mass.g > 0 & mydata$max.life.mo > 0, ]
head(mydata)
```

Add new columns for the logs of mass and gestation:
```{r}
mydata$log_mass <- log(mydata$mass.g)
mydata$log_gest <- log(mydata$gestation.mo)
head(mydata)
```


```{r}
# library(ggplot2) # library already opened previously
main <- ggplot(mydata, aes(log_mass, log_gest)) + geom_point() + theme_bw()
main
class(main)
```

```{r}
sub <- main + geom_rect(data = mydata[1, ], xmin = 0, ymin = -1, xmax = 12, ymax =2.5,
                        fill = "blue", alpha = 0.5)
#* geom_rect = a geometric rectangle
#* alpha = transparency of the rectangle
```

```{r}
sub$layers <- rev(sub$layers) # reorders the layers to draw rectangle below the points
```


```{r}
main + annotation_custom(ggplotGrob(sub), xmin = 0.01, xmax = 5, ymin = 1.5, 
                         ymax = 3.3) + scale_x_continuous(
                           limits = c(0, 12)) + scale_y_continuous(limits = c(-1, 3.3))
# xmin ... ymax: reads the coordinate of the inset within the main panel
# scale_x_continous: the lower and upper limits of the main panel's x axis
# scale_y_continous: the lower and upper limits of the main panel's y axis
```

>-- The main panel now shows just a portion of the plot covered by the rectangle \
>-- the sub panel (showing the blue rectangle and the rest of the entire graph) is insetat the top left of the larger panel

